A few stickleback can transport adaptive alleles to new lakes

### 2019-05-08 Stickleback_Paper.tex
Removed unneccessary files & updated figured top have larger font size

Threespine stickleback provide a striking example of local adaptation to divergent habitats while still being contected by recurrent gene flow. These small fish occur in marine and freshwater habitats throughout the Northern Hemisphere, and in numerous cases smaller freshwater populations have been established ``de novo'' from marine colonists. Previous results show that a similar freshwater phenotype often evolves in independent populations soon after colonization using largely the same standing genetic variation, even in very distant geographic regions. But how many stickleback migrants are needed to capture a sufficient amount of standing variation to fuel adaptation? Here we use genealogical simulations to determine the levels of gene flow that best match observed patterns of allele sharing among habitats in stickleback, and to better understand how gene flow and local adaptation in large metapopulations determine the speed of adaptation and re-use of standing genetic variation. In our simulations we find that repeated adaptation uses a shared set of alleles that are maintained at low frequency by migration-selection balance in ocean populations. This process occurs over a realistic range of intermediate levels of gene flow that match previous empirical population genomic studies in stickleback. Examining these simulations more deeply reveals that low gene flow leads to slow, independent adaption to district habitats, whereas high gene flow leads to significant mutation load but an increased probability of successful population genomic scans for locally adapted alleles. We also find that the original generation of marine individuals that colonized the lakes is most often the sole source of most freshwater adapted alleles. These simulations provide deeper context for existing studies of stickleback evolutionary genomics, and provide guidance for future empirical studies in this model. More broadly, our results support existing theory of local adaptation but extend it by more completely documenting the genealogical history of adaptive alleles in a metapopulation.

@@

In short, the difference between an individual's trait value and the optimum determines that individual's fitness. We chose the difference between optima and strength of stabilizing selection in each habitat so that (a) around 10 (diploid, homozygous) mutations were sufficient to move from one optimum to the other,and (b) well-adapted fish from one habitat would have low, but nonzero, fitness in the other habitat.
% Note that the scale we measure trait on is arbitrary as only the scale of trait values relative to mutation effect size is relevant.

\begin{figure}
    \begin{center}
          \includegraphics[width=0.6\linewidth]{GeographyFigure.pdf}
          \caption{
            \textbf{Diagram of simulated populations:}
            a single, continuous, one-dimensional marine habitat (blue)
            is coupled to randomly mating ``lakes'' at discrete locations with arrows representing migration patterns.
            After an initial period of 100K generations with 25 lakes,
            an additional 25 lakes are added (at the same set of locations) and populated with marine individuals
            to simulate the appearance of newly accessible freshwater habitats colonized by marine stickleback.
            The marine habitat, and each set of 25 lakes, each contain 5,000 individuals at all times.
            }
          \label{fig:Geo}
    \end{center}
\end{figure}

@@

We use SLiM to simulate a Wright--Fisher population with non-overlapping generations and a fixed population size of 5,000 diploid individuals in each habitat. Each generation, the two parents of each new offspring are chosen proportional to their fitness (all individuals are hermaphroditic), and the contributing genomes are produced by Poisson recombinationwith an average of one crossover per chromosome per generation ($10^{-7}$ per locus per generation). Since the total population across \emph{all} 25 lakes is fixed at 5000, and the Wright--Fisher model assumes unrealistic global population regulation, we normalize the fitnesses of each individual such that approximately 200 offspring are generated in each lake, each generation. To do this, we divide fitness values of each freshwater individual by the mean fitness in their lake, so that the mean fitnesses of all lakes are equal before selection happens.

As depicted in Figure~\ref{fig:Geo}, dispersal occurs both locally along the coastline in the marine habitat as well as between the marine habitat and the lakes,
with a lake--ocean migration rate denoted $m$.All individual dispersal events can be thought of as occurring at the juvenile stage in the life cycle of the simulation. 
Each new individual in each habitat has parents from the other habitat with probability $m$ (in which case we call it a ``migrant''),and parents from the same habitat with probability $1-m$.
% how do we get nonmigrants?
The first parent of each nonmigrant individual in the freshwater habitat is chosen from the freshwater habitat proportional to fitness, and a mate is chosen from the same lake as the first, also proportional to fitness. The offspring then lives in the same lake as the parents. Parents for each nonmigrant marine individual are chosen similarly:
first, a single parent is chosen proportionally to fitness in the marine habitat, and then a mate is chosen, also proportionally to fitness but re-weighted by a Gaussian function of the distance separating the two, with standard deviation $1/2$. Concretely, if the first parent is marine individual $i$, then marine individual $j$ is chosen as the mate with probability proportional to $f(x_j) \exp(-2d_{ij}^2)$, where $f(x_j)$ is the fitness of individual $j$ and $d_{ij}$ is the distance between the two locations. 
Finally, each new marine offspring is given a position displaced from the first parent's position by a random Gaussian distance with mean 0 and standard deviation 0.5, and reflected to stay within the habitat range.
% what about migrants?
Parents for each freshwater migrant are chosen in the same way as for nonmigrant marine individuals, and are assigned to the lake nearest to the position of the first marine parent. Similarly, parents for each marine migrant are both chosen from the same lake as before, and the offspring is given a spatial location in the marine habitat at the location of the parent's lake.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{Final_Plots/Fst_Genome_faa_pos.pdf}
          \caption{
        Average $F_{ST}$ in windows of 500Mb between:
                \textbf{(top)} marine habitat and old lakes;
                \textbf{(middle)} marine habitat and new lakes; and
                \textbf{(bottom)} old and new lakes.
                Each plot shows $F_{ST}$ values for a separate simulation,
                with columns corresponding to increasing gene flow from left to right.
        $F_{st}$ is calculated between all marine individuals, and all lakes pooled together.
        %Ten vertical dotted pink lines in each subplot 
                %show regions which have the potential to affect phenotype.
        All locations of pre-existing freshwater adapted alleles have been highlighted by 
        pink dashed vertical lines with an alpha value of 0.3, so darker shades of blue imply more 
        freshwater adapted alleles at that location.
     } \label{fig:Fst}
    \end{center}
\end{figure}

@@

\begin{figure}
    \begin{center}
          \includegraphics{Final_Plots/Freshwater_Alleles.pdf}
          \caption{
            Amount of standing freshwater variation by habitat, across migration rates.
            Each plot counts ``pre-existing freshwater adapted alleles'',
            that are common in the original lakes but rare in the ocean
            (see text for definition).
            \textbf{(A)} Mean number of these alleles per individual.
            \textbf{(B)} Mean percentage of these alleles per individual.
            \textbf{(C)} Total number of these alleles (so, $B = A/C$).
            The number of alleles meeting these conditions changes over the course of the simulation,
            and each plot shows distributions of these values across generations.
            The horizontal axis shows $M$, the mean number of migrants per lake per generation.
        }
        \label{fig:MPFAI}
    \end{center}
\end{figure}

\begin{figure}
    \begin{center}
    \includegraphics[width=0.7\linewidth]{Final_Plots/Allele_Origin_2.pdf}
    \caption{ 
        \textbf{(Origin of adaptive alleles:)}
        Each bar plot shows the origins
        of all trait-affecting alleles above frequency 50\% in at least one new lake,
        classified as
        \textbf{(red)} new mutations,
        \textbf{(blue)} post-colonization migrants,
        \textbf{(green)} ``captured'' from pre-existing lakes, or
        \textbf{(orange)} standing marine variation.
        See Methods for precise definitions of these categories.
    } \label{fig:Origin}
    \end{center}
\end{figure}

@@

Suppose a new allele enters a lake, either by migration or mutation. If, when it is rare but present in $n$ copies, it has fitness advantage $s$ -- i.e., the expected number of copies in the next generation is $(1+s)n$ -- then the probability that it escapes demographic stochasticity to become common in the population is approximately $2s$ \citep{lambert2006probability,haldane1927mathematical(assuming Poisson reproduction, as we roughly have here). If the current population all differ from the optimum trait by $z$, and the allele has effect size $-u$ in heterozygotes,  then the fitness advantage of the allele would be $s(u) = \exp(-\beta((z - u)^2) / \exp( - \beta z^2)) \approx 2 \beta z u$, where in our parameterization, $\beta = 1 / 450$. This tells us two things: (1) the rate of adaptation decreases as the population approaches the optimum, and (2) larger mutations (in the right direction) are more likely to fix.

@@

%\paragraph{Implications for the ``Transporter"--hypothesis}

@@

\begin{figure}
    \begin{center}
          \includegraphics[width=1.0\linewidth]{Final_Plots/Haplo_small.pdf}
          \caption{
            Phenotypes (\textbf{Top}) (effect size and mutation type) and haplotypes \textbf{(Botton)} as a function of spatial position of the individual 
            possessing the genome. The phenotype of each haplotype is, by definition, the sum of the 
            effect sizes of each mutation in the genome. 
        }
          \label{fig:Haplo_Pheno}
    \end{center}
\end{figure}

@@

\begin{figure}
    \begin{center}
          \includegraphics[width=1.0\linewidth]{Final_Plots/Fst_Genome_faa_0_5_500.pdf}
          \caption{
        Average $F_{ST}$ in windows of 500Mb between:
                \textbf{(top)} marine habitat and old lakes;
                \textbf{(middle)} marine habitat and new lakes; and
                \textbf{(bottom)} old and new lakes.
                Each plot shows $F_{ST}$ values for a separate simulation,
                with columns corresponding to increasing gene flow from left to right.
        $F_{st}$ is calculated between all marine individuals with spatial location $\le 5.0$, and the five corresponding lakes.
        All locations of pre-existing freshwater adapted alleles have been highlighted by 
        pink dashed vertical lines with an alpha value of 0.3, so darker shades of blue imply more 
        freshwater adapted alleles at that location.
        }
          \label{fig:fst_5lakes}
    \end{center}
\end{figure}

>>>>

Threespine stickleback provide a striking example of local adaptation to divergent habitats while still being connected by recurrent gene flow. These small fish occur in marine and freshwater habitats throughout the Northern Hemisphere, and in numerous cases smaller freshwater populations have been established ``de novo'' from marine colonists. Previous results show that a similar freshwater phenotype often evolves in independent populations soon after colonization using largely the same standing genetic variation, even in very distant geographic regions. But how many stickleback migrants are needed to capture a sufficient amount of standing variation to fuel adaptation? Here we use genealogical simulations to determine the levels of gene flow that best match observed patterns of allele sharing among habitats in stickleback, and to better understand how gene flow and local adaptation in large metapopulations determine the speed of adaptation and re-use of standing genetic variation. In our simulations we find that repeated adaptation uses a shared set of alleles that are maintained at low frequency by migration-selection balance in ocean populations. This process occurs over a realistic range of intermediate levels of gene flow that match previous empirical population genomic studies in stickleback. Examining these simulations more deeply reveals that low gene flow leads to slow, independent adaption to district habitats, whereas high gene flow leads to significant mutation load but an increased probability of successful population genomic scans for locally adapted alleles. We also find that the original generation of marine individuals that colonized the lakes is most often the sole source of most freshwater adapted alleles. These simulations provide deeper context for existing studies of stickleback evolutionary genomics, and provide guidance for future empirical studies in this model. More broadly, our results support existing theory of local adaptation but extend it by more completely documenting the genealogical history of adaptive alleles in a metapopulation.

@@

In short, the difference between an individual's trait value and the optimum determines that individual's fitness. We chose the difference between optima and strength of stabilizing selection in each habitat so that (a) around 10 (diploid, homozygous) mutations were sufficient to move from one optimum to the other, and (b) well-adapted fish from one habitat would have low, but nonzero, fitness in the other habitat.

\begin{figure}
    \begin{center}
          \includegraphics[width=0.8\linewidth]{Final_Plots/GeographyFigure.pdf}
          \caption{
            \textbf{Diagram of simulated populations:}
            a single, continuous, one-dimensional marine habitat (blue)
            is coupled to randomly mating ``lakes'' at discrete locations with arrows representing migration patterns.
            After an initial period of 100K generations with 25 lakes,
            an additional 25 lakes are added (at the same set of locations) and populated with marine individuals
            to simulate the appearance of newly accessible freshwater habitats colonized by marine stickleback.
            The marine habitat, and each set of 25 lakes, each contain 5,000 individuals at all times.
            }
          \label{fig:Geo}
    \end{center}
\end{figure}

@@

We use SLiM to simulate a Wright--Fisher population with non-overlapping generations and a fixed population size of 5,000 diploid individuals in each habitat. Each generation, the two parents of each new offspring are chosen proportional to their fitness (all individuals are hermaphroditic), and the contributing genomes are produced by Poisson recombination with an average of one crossover per chromosome per generation ($10^{-7}$ per locus per generation). Since the total population across \emph{all} 25 lakes is fixed at 5000, and the Wright--Fisher model assumes unrealistic global population regulation, we normalize the fitnesses of each individual such that approximately 200 offspring are generated in each lake, each generation. To do this, we divide fitness values of each freshwater individual by the mean fitness in their lake, so that the mean fitnesses of all lakes are equal before selection happens.

As depicted in Figure~\ref{fig:Geo}, dispersal occurs both locally along the coastline in the marine habitat as well as between the marine habitat and the lakes,
with a lake--ocean migration rate denoted $m$.All individual dispersal events can be thought of as occurring at the juvenile stage in the life cycle of the simulation. 
Each new individual in each habitat has parents from the other habitat with probability $m$ (in which case we call it a ``migrant''),and parents from the same habitat with probability $1-m$.
The first parent of each nonmigrant individual in the freshwater habitat is chosen from the freshwater habitat proportional to fitness, and a mate is chosen from the same lake as the first, also proportional to fitness. The offspring then lives in the same lake as the parents. Parents for each nonmigrant marine individual are chosen similarly:
first, a single parent is chosen proportionally to fitness in the marine habitat, and then a mate is chosen, also proportionally to fitness but re-weighted by a Gaussian function of the distance separating the two, with standard deviation $1/2$. Concretely, if the first parent is marine individual $i$, then marine individual $j$ is chosen as the mate with probability proportional to $f(x_j) \exp(-2d_{ij}^2)$, where $f(x_j)$ is the fitness of individual $j$ and $d_{ij}$ is the distance between the two locations. 
Finally, each new marine offspring is given a position displaced from the first parent's position by a random Gaussian distance with mean 0 and standard deviation 0.5, and reflected to stay within the habitat range.
Parents for each freshwater migrant are chosen in the same way as for nonmigrant marine individuals, and are assigned to the lake nearest to the position of the first marine parent. Similarly, parents for each marine migrant are both chosen from the same lake as before, and the offspring is given a spatial location in the marine habitat at the location of the parent's lake.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{Final_Plots/Fst_Genome_faa.pdf}
          \caption{
        Average $F_{ST}$ in windows of 500Mb between:
                \textbf{(top)} marine habitat and old lakes;
                \textbf{(middle)} marine habitat and new lakes; and
                \textbf{(bottom)} old and new lakes.
                Each plot shows $F_{ST}$ values for a separate simulation,
                with columns corresponding to increasing gene flow from left to right.
        $F_{st}$ is calculated between all marine individuals, and all lakes pooled together.
        %Ten vertical dotted pink lines in each subplot 
                %show regions which have the potential to affect phenotype.
        All locations of pre-existing freshwater adapted alleles have been highlighted by 
        pink dashed vertical lines with an alpha value of 0.3, so darker shades of pink imply more 
        freshwater adapted alleles at that location.
     } \label{fig:Fst}
    \end{center}
\end{figure}

@@

\begin{figure}
    \begin{center}
          \includegraphics{Final_Plots/Freshwater_Alleles.pdf}
          \caption{
            Amount of standing freshwater variation by habitat, across migration rates.
            Each plot counts ``pre-existing freshwater adapted alleles'',
            that are common in the original lakes but rare in the ocean
            (see text for definition).
            \textbf{(A)} Mean number of these alleles per individual.
            \textbf{(B)} Mean percentage of these alleles per individual.
            \textbf{(C)} Total number of these alleles (so, $B = A/C$).
            The number of alleles meeting these conditions changes over the course of the simulation,
            and each plot shows distributions of these values throughout each simulations.
            The horizontal axis shows $M$, the mean number of migrants per lake per generation.}
        \label{fig:MPFAI}
    \end{center}
\end{figure}

\begin{figure}
    \begin{center}
    \includegraphics{Final_Plots/Allele_Origin_2.pdf}
    \caption{ 
        \textbf{(Origin of adaptive alleles:)}
        Each bar plot shows the origins
        of all trait-affecting alleles above frequency 50\% in at least one new lake,
        classified as
        \textbf{(red)} new mutations,
        \textbf{(blue)} post-colonization migrants,
        \textbf{(green)} ``captured'' from pre-existing lakes, or
        \textbf{(orange)} standing marine variation.
        See Methods for precise definitions of these categories.
    } \label{fig:Origin}
    \end{center}
\end{figure}

@@

Suppose a new allele enters a lake, either by migration or mutation. If, when it is rare but present in $n$ copies, it has fitness advantage $s$ -- i.e., the expected number of copies in the next generation is $(1+s)n$ -- then the probability that it escapes demographic stochasticity to become common in the population is approximately $2s$ \citep{lambert2006probability,haldane1927mathematical} (assuming Poisson reproduction, as we roughly have here). If the current population all differ from the optimum trait by $z$, and the allele has effect size $-u$ in heterozygotes,  then the fitness advantage of the allele would be $s(u) = \exp(-\beta((z - u)^2) / \exp( - \beta z^2)) \approx 2 \beta z u$, where in our parameterization, $\beta = 1 / 450$. This tells us two things: (1) the rate of adaptation decreases as the population approaches the optimum, and (2) larger mutations (in the right direction) are more likely to fix.

@@

@@

\begin{comment}
\begin{figure}
    \begin{center}
          \includegraphics[width=1.0\linewidth]{Final_Plots/Haplo_small.pdf}
          \caption{
            Phenotypes (\textbf{Top}) (effect size and mutation type) and haplotypes \textbf{(Botton)} as a function of spatial position of the individual 
            possessing the genome. The phenotype of each haplotype is, by definition, the sum of the 
            effect sizes of each mutation in the genome. 
        }
          \label{fig:Haplo_Pheno}
    \end{center}
\end{figure}
\end{comment}

@@

\begin{figure}
    \begin{center}
          \includegraphics[width=1.0\linewidth]{Final_Plots/Fst_Genome_faa0_5_500.pdf}
          \caption{
        Average $F_{ST}$ in windows of 500Mb between:
                \textbf{(top)} marine habitat and old lakes;
                \textbf{(middle)} marine habitat and new lakes; and
                \textbf{(bottom)} old and new lakes.
                Each plot shows $F_{ST}$ values for a separate simulation,
                with columns corresponding to increasing gene flow from left to right.
        $F_{st}$ is calculated between all marine individuals with spatial location $\le 5.0$, and the five corresponding lakes.
        All locations of pre-existing freshwater adapted alleles have been highlighted by 
        pink dashed vertical lines with an alpha value of 0.3, so darker shades of blue imply more 
        freshwater adapted alleles at that location.
        }
          \label{fig:fst_5lakes}
    \end{center}
\end{figure}
